\thetitle

### 2020-07-07 writing/challenges.tex
prose touch ups

The most serious challenge for comparing blame assignment strategies concerns 
the nature of this research. Unlike ordinary
work on programming languages, the question seems to call for empirical research
similar to the one in the human-computer interaction area. At the same time, a
significant evaluation result demands a large number of cases yielding a serious
amount of data. But, as~\citet{lksfd-popl-2020} already demonstrated, the way
around this dilemma is to postulate the equivalent of a {\it homo economicus\/}
for programming languages---the {\em rational programmer\/}---and to implement
this rational programmer as an algorithm.  As indicated in the introduction, in
the context of gradual typing, the rational programmer translates the
Wadler-Findler slogan into a debugging method, searching for the source of the
impedance mismatch in an incremental fashion. Measuring this simulated
programmer's behavior on a large number of inputs is the equivalent of
collecting empirical data in a truly human-focused discipline.

One derivative challenge of the first one is the problem of simulating the
programmer's mistakes or, more precisely, the origin of impedance
mismatches. Before a programmer can try to find mistakes, mistakes have to
be made. Technically the problem is the lack of a sufficiently large
available corpus of type errors for a meaningful experiment.
Following~\citet{lksfd-popl-2020} again, the solution is to use mutation
analysis~\cite{lipton1971fault, demillo1978hints, jia2011analysis} to
create mistakes in a synthetic manner. This process yields a large number
of buggy programs whose mixed-typed versions are suitable subjects for our
experiment.  The mutation analysis of Lazarek et al, is, however,
completely useless for creating impedance mismatches in a gradually
typed context.

@@

Another derivative challenge of the first one is the problem of mistakes in type
annotations themselves.  Gradual typing enables programmers to move an untyped
code base into the typed realm on an incremental basis. The retroactive addition
of a specific type to a component, even a library, may result from a
misunderstanding of the code and may thus be a mistake itself. Empirical
evidence suggests that this scenario is quite common~\cite{sta-nt-base-types,
incorrect-ts, wmwz-ecoop-2017}, and hence must be addressed by any evaluation
method. Conversely, the method not only demands mutators for code that cause
good impedance mismatches, it also calls for a set of mutators that ``break''
type annotations without breaking the code itself.

The second big challenge is due to the implementation disparity of gradual
type systems. While the various approaches are well understood in theory,
they come with vastly different type systems, accommodate seemingly
incompatible language features, and apply to radically different coding
styles in the untyped world.  To eliminate these differences as variables
of a comparison, the application of an evaluation method to different
gradual typing systems must use the same test programs and the same type
system. Otherwise, results for different gradual typing systems are not
comparable.  For instance, the case study presented here builds on
Greenman and Felleisen's transient variant for functional Typed
Racket~\cite{gf-icfp-2018} to obtain a transient semantics that is
comparable with Typed Racket's default natural one. The revised transient
variant accommodates many more linguistic features and calls library code
in a in a way that is faithful to the transient
semantics~\cite{vss-popl-2017}.

@@

Making these strategies comparable is {\em not\/} an issue of language
implementation but one about how the rational programmer behaves with regard to
these messages. Specifically, the rational programmer must come with \emph{modes}
that represent the alternative ways in which a programmer may interpret blame. 
For instance, in transient, one mode has the competent programmer
pick the oldest element of a blame set as the next component to add types. After
all, this corresponds to the earliest point in the program's evaluation that can
discover an impedance mismatch.  Another mode does not distinguish between the elements of the
blame set and requires that the rational programmer types all blamed components
at once. Of course, a step of the rational programmer in the latter mode
is not as laborious as one in the former. Hence, an evaluation method should take 
into account the effort of the rational programmer when comparing 
different modes.

@@

These challenges expose the subtlety of evaluating blame strategies but
also suggest a recipe for doing so correctly:

>>>>

The most serious challenge for comparing blame assignment strategies concerns 
the nature of this research. Unlike ordinary
work on programming languages, the question seems to call for empirical research
similar to the one in the human-computer interaction area. At the same time, a
significant evaluation result demands a large number of cases yielding a serious
amount of data. But, as~\citet{lksfd-popl-2020} already demonstrated, the way
around this dilemma is to postulate the equivalent of a {\it homo economicus\/}
for programming languages---the {\em rational programmer\/}---and to implement
this rational programmer as an algorithm.  As indicated in the introduction, in
the context of gradual typing, the rational programmer translates the
Wadler-Findler slogan into a debugging method, searching for the source of the
impedance mismatch in an incremental fashion. Measuring this simulated
programmer's behavior on a large number of inputs then yields relevant 
data, similar to collecting empirical data in a human-focused discipline.

One derivative challenge of the first one is the problem of simulating the
programmer's mistakes or, more precisely, the origin of impedance
mismatches. Before a programmer can try to find mistakes, mistakes have to
be made. Technically the problem is the lack of a sufficiently large
available corpus of type errors for a meaningful experiment.
Following~\citet{lksfd-popl-2020} again, the solution is to use mutation
analysis~\cite{lipton1971fault, demillo1978hints, jia2011analysis} to
create mistakes in a synthetic manner. This process yields a large number
of buggy programs whose mixed-typed versions are suitable subjects for our
experiment.  The mutation analysis of Lazarek et al, is, however,
completely useless for creating impedance mismatches.

@@

Another derivative challenge is the problem of mistakes in type
annotations themselves.  Gradual typing enables programmers to move an untyped
code base into the typed realm on an incremental basis. The retroactive addition
of a specific type to a component, even a library, may result from a
misunderstanding of the code and may thus be a mistake itself. Anecdotal 
evidence suggests that this scenario is quite common~\cite{sta-nt-base-types,
incorrect-ts, wmwz-ecoop-2017}, and hence must be addressed by any evaluation
method. Conversely, the method not only demands mutators for code that cause
good impedance mismatches, it also calls for a set of mutators that ``break''
type annotations without breaking the code itself.

The second big challenge is the disparity of gradually
typed languages. While the approaches are well understood in theory,
they come with vastly different type systems, accommodate seemingly
incompatible language features, and apply to radically different coding
styles in the untyped world.  To eliminate these differences as variables
of a comparison, the application of an evaluation method to different
gradual typing systems must use the same test programs and the same type
system. Otherwise, results for different gradual typing systems are not
comparable.  For instance, the case study presented here builds on
Greenman and Felleisen's transient variant for functional Typed
Racket~\cite{gf-icfp-2018} to obtain a transient semantics that is
comparable with Typed Racket's default natural one. The revised transient
variant accommodates many more linguistic features and library calls
in a way that is faithful to the transient
semantics~\cite{vss-popl-2017}.

@@

Making these strategies comparable is {\em not\/} an issue of language
implementation but the behavior of the rational programmer.
Specifically, the rational programmer must come with \emph{modes}
that represent the alternative interpretations of blame. 
For instance, a programmer using transient may
pick the oldest element of a blame set as the next component to add types because
it corresponds to the earliest point in the program's evaluation that can
discover an impedance mismatch.  Another mode does not distinguish between the elements of the
blame set and has the rational programmer type all blamed components
at once. Because this second kind of step 
is more laborious than the first one, a comparison must take 
into account the rational programmer's effort.

@@

These challenges expose the subtlety of evaluating blame strategies but
also suggest a recipe:

%% for doing so correctly
%% MF: what is correct about this recipe?

### 2020-07-07 writing/introduction.tex
prose touch ups

The contrast between academic and industrial systems raises the question whether
\begin{quote}
 \it
 blame assignment adds any value to a gradually typed language, \\
 especially for the benefit of the working programmer.
\end{quote}
Given the prominence of blame in academia and its complete absence in
industrial systems, it comes as an even bigger surprise that the research
literature and the industrial blog world do not present or discuss any
possible answers.  Instead, when language designers decide to look into the
question, they seem to answer it one way or another without any scientific
justification. Most of the time though they ignore the question altogether. The
root of the issue is that the research community simply does not have a
method for evaluating such questions, and given the question's practical
relevance, this a significant problem.

This paper's first contribution is an automated method for evaluating the
effectiveness of blame assignment strategies in the gradual typing world
(sections ~\ref{sec:landscape},~\ref{sec:mutate},~\ref{sec:rational} and \ref{sec:sample}). 
Roughly speaking, the method, which is
loosely based on the work of Lazarek et al.~\cite{lksfd-popl-2020},
simulates a programmer who follows the Wadler-Findler slogan quoted above.
Such a \emph{rational programmer} can find impedance-mismatch
bugs by adding types to a partially typed program. That is, when the
run-time system signals an impedance mismatch, a good blame message helps
the programmer infer where to add an additional type specification. Since
components that pass the type checker cannot cause impedance mismatches;
the type checker discovers these mismatches at compile time and points
directly to the faulty component. Otherwise, if the
additional type does {\em not\/} cause a static type error, the program
 is run again to signal another violation. Then the programmer can heed
blame one more time to add yet another type specification. Counting the number of
steps it takes to discover the mistake in this manner is a measure of the effectiveness
of the blame assignment strategy of the language.

@@

The results of the experiment (section~\ref{sec:discussion}) validate the
conjecture behind the work of theoreticians. A good blame assignment
strategy
greatly shortens the search for impedance mismatches between the specified types
and the behavior of untyped components. Even better, the evaluation method clearly
identifies the wrapper-based blame tracking of Typed Racket as the vastly superior
academic system. Vitousek et al. lose one more time. Of course, a detailed
discussion must acknowledge the threats to validity of these results.

>>>>

The contrast between academic and industrial systems raises the question 
\begin{quote}
 \it
 whether blame assignment adds any value to a gradually typed language, \\
 especially for the benefit of the working programmer.
\end{quote}
Given the prominence of blame in academia and its complete absence in
industrial systems, it comes as an even bigger surprise that the research
literature and the industrial blog world do not present or discuss any
possible answers.  Instead, when language designers decide to look into the
question, they seem to answer it one way or another without any scientific
justification. As a matter of fact, the research community simply does not have a
method for evaluating blame, and given the question's practical
relevance, this a significant problem.

This paper's first contribution is an automated method for evaluating the
effectiveness of blame assignment strategies in the gradual typing world.
%% (sections ~\ref{sec:landscape},~\ref{sec:mutate},~\ref{sec:rational} and \ref{sec:sample}). 
%% MF: says you can't simulate a roadmap by listing all sections.
%% MF: and for heaven's sake lookup the use of ~ in TeX
Roughly speaking, the method, which is
loosely based on the work of Lazarek et al.~\cite{lksfd-popl-2020},
simulates a programmer who follows the Wadler-Findler slogan quoted above.
Such a \emph{rational programmer} can find impedance-mismatch
bugs by adding types to a partially typed program (section~\ref{sec:landscape}). That is, when the
run-time system signals an impedance mismatch, a good blame message helps
the programmer infer where to add an additional type specification. Since
components that pass the type checker cannot cause impedance mismatches;
the type checker discovers these mismatches at compile time and points
directly to the faulty component. Otherwise, if the
additional type does {\em not\/} cause a static type error, the program
 is run again to signal another violation. Then the programmer can heed
blame one more time to add yet another type specification. Weighing the number of
steps it takes to discover the mistake in this manner is a measure of the effectiveness
of the blame assignment strategy of the language.

@@

A direct interpretation of the results (section~\ref{sec:discussion}) is that
they validate the conjecture behind the work of theoreticians. A good blame assignment
strategy
greatly shortens the search for impedance mismatches between the specified types
and the behavior of untyped components. Even better, the evaluation method clearly
identifies the wrapper-based blame tracking of Typed Racket as the vastly superior
academic system. Vitousek et al. lose one more time. Of course, a detailed
discussion must acknowledge the threats to validity of these results.
